---
title: "Homework 1"
author: "Kevin Sanford"
date: "10-16-2024"
format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
    code-line-numbers: true
error: true
editor: 
  markdown: 
    wrap: 72
---

# Instructions

-   This problem set contains 2 exercises in total.
-   **You may only use functions from Base R to complete the
    exercises**. Email Dr. Landeros (`alandero@ucr.edu`) if you have any
    doubts.
-   Check the Canvas Discussion section for updates.

# Background

Many problems in statistics are challenging precisely because at least
one key distribution is either unknown or unavailable. Otherwise, many
statistics and quantities of interest easy to calculate. We exploit this
fact in constructing a data structure for *approximating* various
statistics of a data stream.

These two exercises problem guides you in constructing an *expanding
histogram* that you can leverage in calculating quantiles. The idea is
simple:

1.  A user specifies the number of bins $b$ which **must** be even. This
    information is used to construct $b-1$ *left-closed* bins and $1$
    *closed* bin (the rightmost one):

    ``` text
    [a, b), [b, c), [c, d), [d, e]
    ```

    This numbers $a$, $b$, $c$, $d$, and $e$ are the *edges*.

2.  Now suppose we start with the following bin counts:

    ``` text
    |1|2|5|3|
    ```

    where the extreme edges are $a$ and $e$.

3.  We observe a data point $y$ and so we must update the histogram.
    There are 3 cases to consider:

    -   **Case:** $a \le y \le b$: The point lies inside the current
        limits of the histogram, so we simply find the bin it belongs
        to.

        ``` text
           *
           ↓
        |1|2|5|3|   --->   |1|3|5|3|
        ```

    -   **Case:** $y > b$: The point lies outside (to the right) of bin
        edges.

        ``` text
        |1|2|5|3|       *
        ```

        We expand each bin to the right by doubling the edges. Adjacent
        bins merge with their neighbors to the right. Multiple
        expansions (doublings) may be required to cover the data:

        ``` text
        |1|2|5|3|       *
         \ / \ / \ / \ /
        | 3 | 8 | 0 | 1 |
        ```

    -   **Case:** $y < a$: This case is symmetric to the previous case,
        except the histogram expands to the left.

4.  Given our histogram and a probability $p \in [0,1]$, we perform a
    *search* to find $x$ such that $\Pr(X \le x) \le p$. The value $x$
    is an approximation of the target quantile.

This exercise was inspired by Josh Day's and Hua Zhou's software
package, OnlineStats.jl:

> Day J, Zhou H (2020). OnlineStats.jl: A Julia package for statistics
> on data streams. *Journal of Open Source Software*, 5(46), 1816, doi:
> [10.21105/joss.01816](https://doi.org/10.21105/joss.01816)

A few additional remarks:

-   While we are not aiming for high performance within R, you should be
    mindful of how you structure your code.

-   We will create our own S3 class to represent an expanding histogram.
    **We did not discuss S3 classes in lecture, so you do not need to
    worry about these details.** The exercises will give you a sense of
    how object-oriented programming works in R's S3 system, but all the
    work has been done for you.

-   Problem 1 is relatively more challenging than Problem 2.

-   Problem 2 implements statistics for the expanding histogram. It is
    written in a way that allows us to grade it independently of Problem
    1, but you can test out the functions on the object you created.

# Problem 1: An expanding histogram

## Representing a histogram

The `hist()` function in R is useful for plotting. In the background, it
creates a `"histogram"` class object to fit a histogram to data. We can
access this object by passing the `plot = FALSE` option to `hist()`.

**Modify the code chunk below to do the following**:

1.  Create a `"histogram"` object fit to the given data, `xdata`, as
    described above.
2.  Save this object to a variable called `myhist`.
3.  Display `myhist` in your report.
4.  *Below the code chunk*, write down the number of *fields* for this
    object (do not count the `attr(,"class")` which appears).

Study these fields, and consult `help("hist")` if you need
clarification.

```{r}
set.seed(206)
xdata <- rnorm(100)
myhist <- hist(xdata, plot = FALSE)
plot(myhist)
```

6 fields exist for this object.

## Writing our own representation

We begin by defining a data structure to represent the expanding
histogram. **Complete the function below by filling in the following
details**

-   The function should accept a single argument, `N`, which represents
    the number of bins. \*Remember than `N` should be even.

-   Create a `list` object named `h` to represent the expanding
    histogram.

-   The `list` should have the following fields:

    -   `n`, the number of observations used to estimate the histogram.

    -   `edges`, a vector of length `N+1` representing the bin edges.
        Initialize to be `0` everywhere.

    -   `counts`, a vector of length `N` representing the counts in each
        bin. Initialize to be `0` everywhere.

```{r}
init_expanding_hist <- function (N) {
  edges <- rep(0, N+1)
  counts <- rep(0, N)
  h <- list(n = 0, edges = edges, counts = counts)
  # Make it an S3 class
  class(h) <- "expanding_hist"
  return(h)
}
```

## Finding the correct bin

Now that we've outlined an interface, let's begin by handling placement
into bins.

Write a function called `find_bin_index()` that identifies the bin which
covers a new observation `y`. *This function will not do any expanding,
it only retrieves a number indicating a bin.*

-   It should accept two inputs: `edges` and `y`.
-   First, handle all the exceptional cases. Two examples
    (non-exhaustive):
    -   If `y` lies to the left of the histogram, the bin index is `0`
        because it is outside.
    -   If `y` is equal to the rightmost *edge*, it is inside the
        histogram. What is the bin number?
-   Lastly, use `findInterval()` to recover the bin index within the
    histogram.
-   This function should return a single number, the bin index.

```{r}
find_bin_index <- function(edges, y) {
  if (y < edges[1]) {
    return (0)
  }
  if (y > edges[length(edges)]) {
    return (length(edges))
  }
  bin_number <- findInterval(y, edges, rightmost.closed = TRUE)
  return(bin_number)
}

```

## Merging Left

Next we handle the case in which the histogram had to be expanded to
cover a new observation.

Write a function called `merge_left()` which merges bin `counts`,
*starting from the rightmost bin and moving towards the leftmost one*.
This procedure requires three pieces of information:

-   `counts`, the bin counts.
-   `K`, the number of times the histogram was expanded.
-   `M`, the middle bin index ($M = N / 2$ since $N$ is even).
-   Return the modified counts.

**Hints**:

-   Assume `K` and `M` are given to you and that they are correctly
    computed.
-   What does `K` tell you about the number of required merges? What
    about `M`?
-   This procedure requires two loops.

```{r}
merge_left <- function(counts, K, M) {
  # define pointer to keep track of elements to merge
  p <- length(counts)
  # define stopping point for merges
  if (K > 0) {
    s <- M + floor(M/(2^K))
  }
  else{
    s <- M
  }
  # merge bins from the right and rewrite into counts
  for (i in length(counts):(s+1)) {
    counts[i] <- counts[p] + counts[p-1]
    # move pointer
    p <- p - 2
  }
  for (i in s:1) {
    counts[i] <- 0
  }
  return(counts)
}

```

## Merging Right

Now we handle the other expansion case.

Write a function called `merge_right()` which merges bin `counts`,
*starting from the leftmost bin and moving towards the rightmost one*.

```{r}
merge_right <- function(counts, K, M) {
  # define pointer to keep track of elements to merge
  p <- 1
  # define stopping point for merges
  s <- ceiling(M/(2^K))
  # merge bins from the left and rewrite into counts
  for (i in 1:s) {
    counts[i] <- counts[p] + counts[p+1]
    # move pointer
    p <- p + 2
  }
  for (i in (s+1):length(counts)) {
    counts[i] <- 0
  }
  return(counts)
}

```

## Expanding the histogram

Next, we combine the merge functions to handle the expansion logic.

Finish implementing the `expand()` function below, using the comments as
a guide.

**Hints**:

-   This function must check which direction to expand in.
-   Expansion should modify the `edges` and `counts` fields only.
-   We may not need to modify `h` at all. When is this the case?
-   Respect the interface: you do not need additional inputs.
-   Note an important invariant: If we expand, exactly one of the
    extreme edges remains fixed.
-   If the current width of the histogram is $0$, then use the new width
    implied by the update due to the new observation `y`.

```{r}
# Inputs:
#  h, an expanding_hist object
#  y, a new observation
# Output:
#  h, the expanding_hist object
expand <- function (h, y) {
  # Retrieve extreme edges
  a <- h$edges[1]
  b <- h$edges[length(h$edges)]
  
  # if the histogram does not need to be expanded, do nothing
  if (y >= a && y <= b) {
    return (h)
  }
  
  # Compute the width of the histogram
  w <- b - a
  # Calculate the middle bin index, M
  M <- length(h$counts)/2
  # initialize K
  K <- 0
  # calculate new bin width
  binw <- 2*(b-a)/length(h$counts)
  E <- length(h$edges)

  # Expand to the right by finding K such that
  # y <= a + 2^K * w
  if (y > b) {
    while (y > a + (2^K)*w) {
      # expand histogram to the right
      h$counts <- merge_right(h$counts, K, M)
      # apply new bin width
      for (i in 2:E) {
        h$edges[i] <- h$edges[1] + (i-1)*binw
      }
      # double bin width for next iteration
      binw <- binw*2
      K <- K + 1
    }
  }
  # Expand to the left by finding K such that
  # y >= b - 2^K * w
  if (y < a) {
    while (y < b - (2^K)*w) {
      # expand histogram to the left
      h$counts <- merge_left(h$counts, K, M)
      # apply new bin width
      for (i in (E-1):1) {
        h$edges[i] <- h$edges[E] - (E-i)*binw
      }
      # double bin width for next iteration
      binw <- binw*2
      K <- K + 1
    }
  }

  return(h)
}
```

## Putting it all together

Fill in the details for the function below to complete the `fit()`
function.

```{r}
fit <- function (h, y) {
  stopifnot(is.finite(y))
  h$n <- h$n + 1
  N <- length(h$counts)

  if (h$n == 1) {
    # Case: Starting with a fresh histogram.
    # There is only 1 bin, so all edges are equal to each other.
    h$edges <- rep(y,length(h$edges))
    h$counts[N] <- h$counts[N] + 1
  } else if (all(h$edges == h$edges[1])) {
    # Case: We've never expanded before.
    a <- h$edges[1]
    if (y == a) {
      # Don't need to expand.
      h$counts[N] <- h$counts[N] + 1
    } else if (y < a) {
      # Expand once to the left.
      h$edges[1] <- y
      # set equal bin widths between a and y
      h$edges <- seq(y, a, length = length(h$edges))
      h$counts[1] <- 1
    } else {
      # Expand once to the right.
      h$edges[N] <- y
      # set equal bin widths between a and y
      h$edges <- seq(a, y, length = length(h$edges))
      # new value took the count from the old value
      # this statement recounts the old value
      h$counts[1] <- 1
    }
  } else {
    # expand before we update the counts.
    h <- expand(h,y)
    # find where y belongs in the expanded histogram
    yi <- find_bin_index(h$edges,y)
    h$counts[yi] <- h$counts[yi] + 1
  }

  return(h)
}
```

Test your implementation on the following test case. **Do not modify the
seed or settings.**

**Expected output**:

-   $n = 1000$
-   $\min(\text{Edges}) \approx -2.438$
-   $\max(\text{Edges}) \approx 13.258$
-   $\text{Counts} = {0, 11, 221, 558, 201, 9, 0, 0, 0, 0}$

```{r}
#| label: test-expanding-hist
set.seed(1903)
y <- rnorm(1000, mean = 3, sd = 1.016)
h <- init_expanding_hist(10)
h

for (yi in y) {
  h <- fit(h, yi)
}
h
```

# Problem 2: Evaluating sample statistics

Now we implement functions to calculate descriptive statistics using our
empirical estimate of a distribution. This problem provides some code so
you can test your histogram implementation, but it is graded
independently of Problem 1.

Using `seed = 1903` in code chunk @test-expanding-hist, we have the
following:

| $n$  | $N$  | $\mu_{n}$ | $s_{n}^{2}$ | $Q_{0.5}$ |
|------|------|-----------|-------------|-----------|
| 1000 | 10   | 3.017902  | 1.544215    | 3.016752  |
| 1000 | 100  | 3.004404  | 1.042367    | 3.009793  |
| 1000 | 1000 | 3.000385  | 1.017131    | 2.984907  |

## Midpoint interpolation

Each of the histogram bins represents a range of values. We will
calculate statistics based on the *midpoints* of each implied interval.
For example, the bins $$
\{[a, b), [b,c), [c, d]\}
$$ have midpoints $(a+b)/2$, $(b+c)/2$, and $(c+d)/2$.

Write a function to compute the $N$ midpoints of a length $N+1$ vector
`x`, called `compute_midpoints()`. You may assume `x` is sorted.

**Do not modify here**

```{r}
# Declare a new generic
midpoints <- function (x) {
  UseMethod("midpoints")
}

# Default behavior: assume input is a vector
midpoints.default <- function (x) {
  compute_midpoints(x)
}

# S3 dispatch for expanding_hist
midpoints.expanding_hist <- function (h) {
  return(midpoints(h$edges))
}
```

**Your implementation**

```{r}
compute_midpoints <- function(x) {
  M <- x[1:(length(x)-1)]
  M <- (M + x[2:length(x)])/2 
  return (M)
}
```

## Mean

Given a collection of histogram midpoints `x` and frequencies `w`,
compute the weighted mean of `x`. Call your function `compute_wmean()`.

::: callout-tip
### Hint

Think carefully about what *is* and *isn't* an observation in computing
this statistic.
:::

**Do not modify here**

```{r}
mean.expanding_hist <- function (h, ...) {
  compute_wmean(midpoints(h), h$counts)
}
```

**Your implementation**

```{r}
compute_wmean <- function(x, w) {
  mu <- sum(x*w)/sum(w)
  return (mu)
}
```

## Variance

Given a collection of histogram midpoints `x` and frequencies `w`,
compute the weighted variance of `x`. Call your function
`compute_wvar()`.

::: callout-tip
### Hint

There is an opportunity to cleverly reuse your `compute_mean()` function
here.
:::

**Do not modify here**

```{r}
var <- function (x, y = NULL, na.rm = FALSE, use) { UseMethod("var") }

registerS3method("var", "default", stats::var)

var.expanding_hist <- function (h, ...) {
  compute_wvar(midpoints(h), h$counts)
}
```

**Your implementation**

```{r}
compute_wvar <- function(x, w) {
  # retrieve weighted mean
  mu <- compute_wmean(x, w)
  wvar <- sum(w*(x-mu)^2)/sum(w)
  return(wvar)
}
```

## Quantiles

Given a collection of histogram midpoints `x`, frequencies `w`, and
probabilities `p`, compute the weighted quantiles of `x`. Call your
function `compute_wquantile()`.

::: callout-tip
### Hint

You may use the `weighted.quantile()` function from the
[descriptio](https://cran.r-project.org/web/packages/descriptio/index.html)
package.
:::

**Do not modify here**

```{r}
quantile.expanding_hist <- function (h, probs, ...) {
  compute_wquantile(midpoints(h), h$counts, probs)
}
```

**Your implementation**

```{r}
library(descriptio)
compute_wquantile <- function(x, w, p) {
  return (descriptio::weighted.quantile(x, w, p))
}
```

